\documentclass[12pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[margin=2.5cm]{geometry}
\usepackage{enumitem}
\usepackage{fancyhdr}
\usepackage{titlesec}
\usepackage{xcolor}
\usepackage{tabularx}
\usepackage{multicol}
\usepackage{array}
\usepackage{tcolorbox}
\usepackage[normalem]{ulem}

% Header and footer
\pagestyle{fancy}
\fancyhf{}
\setlength{\headheight}{15pt}
\lhead{Answer Key}
\rhead{PAST PERFECT CONTINUOUS}
\cfoot{\thepage}

% Section formatting
\titleformat{\section}
{\Large\bfseries\color{green!70!black}}
{\thesection}{1em}{}

\titleformat{\subsection}
{\large\bfseries\color{green!50!black}}
{\thesubsection}{1em}{}

% Custom colors
\definecolor{correct}{RGB}{0, 128, 0}
\definecolor{noteblue}{RGB}{0, 80, 160}
\definecolor{easygreen}{RGB}{34, 139, 34}
\definecolor{mediumyellow}{RGB}{255, 165, 0}
\definecolor{hardred}{RGB}{178, 34, 34}

% Custom tcolorbox for answers
\newtcolorbox{answerbox}[1][]{
    colback=green!5!white,
    colframe=correct,
    fonttitle=\bfseries,
    title=#1
}

\newtcolorbox{notebox}[1][]{
    colback=blue!5!white,
    colframe=noteblue,
    fonttitle=\bfseries,
    title=#1
}

\begin{document}

% ============================================================
% TITLE PAGE
% ============================================================

\begin{center}
    {\Huge\bfseries Answer Key}\\[0.5cm]
    {\Large Past Perfect Continuous}\\[0.3cm]
    {\large (I had been doing)}\\[0.5cm]
    \rule{\textwidth}{0.4pt}\\[0.3cm]
    {\large English Grammar in Use — Unit 16}\\[0.2cm]
    {\large 18 Exercises • Complete Solutions with Explanations}\\[0.2cm]
    {\large Includes Exam Practice: IELTS • Cambridge • TOEFL}\\[0.2cm]
    \today
\end{center}

\vspace{0.5cm}

% ============================================================
% SCORING SUMMARY
% ============================================================

\section*{Scoring Summary}

\begin{center}
\begin{tabular}{|l|c|}
\hline
\textbf{Section} & \textbf{Total Points} \\
\hline
EASY (Exercises 1-4) & 36 points \\
\hline
MEDIUM (Exercises 5-8) & 44 points \\
\hline
HARD (Exercises 9-12) & 35 points \\
\hline
REAL-LIFE \& EXAM (Exercises 13-18) & 55 points \\
\hline
\textbf{GRAND TOTAL} & \textbf{170 points} \\
\hline
\end{tabular}
\end{center}

\newpage

% ============================================================
% EASY SECTION ANSWERS
% ============================================================

\begin{center}
{\Large\textcolor{easygreen}{\textbf{— EASY EXERCISES —}}}
\end{center}

\vspace{0.3cm}

% ============================================================
\section{Exercise 1: Form the Past Perfect Continuous (10 points)}

\begin{answerbox}[Answers]
\begin{enumerate}[label=\arabic*.]
    \item He \textbf{\textcolor{correct}{had been working}} all day.
    \item It \textbf{\textcolor{correct}{had been raining}} for hours.
    \item They \textbf{\textcolor{correct}{had been waiting}} since 3 o'clock.
    \item She \textbf{\textcolor{correct}{had been studying}} English for two years.
    \item We \textbf{\textcolor{correct}{had been living}} in London before we moved.
    \item I \textbf{\textcolor{correct}{had been running}}, so I was out of breath.
    \item He \textbf{\textcolor{correct}{hadn't been sleeping}} well lately.
    \item They \textbf{\textcolor{correct}{had been traveling}} for six months.
    \item She \textbf{\textcolor{correct}{had been sitting}} in the same position for hours.
    \item The cat \textbf{\textcolor{correct}{had been lying}} in the sun all afternoon.
\end{enumerate}
\end{answerbox}

\begin{notebox}[Spelling Notes]
\begin{itemize}
    \item \textbf{sit → sitting} (double the consonant: CVC pattern)
    \item \textbf{lie → lying} (change -ie to -y before -ing)
    \item \textbf{travel → traveling/travelling} (both spellings accepted)
\end{itemize}
\end{notebox}

\textbf{Scoring:} 1 point per correct answer. Total: 10 points.

\vspace{0.5cm}

% ============================================================
\section{Exercise 2: Make Questions (8 points)}

\begin{answerbox}[Answers]
\begin{enumerate}[label=\arabic*.]
    \item \textbf{\textcolor{correct}{What had she been doing}} before you arrived?
    \item \textbf{\textcolor{correct}{How long had they been living}} in Paris before they moved to London?
    \item \textbf{\textcolor{correct}{Had it been raining}} before you left the house?
    \item \textbf{\textcolor{correct}{Why had he been crying}} when you found him?
    \item \textbf{\textcolor{correct}{How long had you been studying}} before you passed the exam?
    \item \textbf{\textcolor{correct}{Where had they been staying}} before they found an apartment?
    \item \textbf{\textcolor{correct}{Who had she been talking}} to on the phone for so long?
    \item \textbf{\textcolor{correct}{Had the children been playing}} outside before it started raining?
\end{enumerate}
\end{answerbox}

\begin{notebox}[Question Formation]
\textbf{Structure:} (Question word) + \textbf{had} + subject + \textbf{been} + verb-\textbf{ing}?
\begin{itemize}
    \item Yes/No questions: \textit{Had you been waiting...?}
    \item Wh- questions: \textit{How long had you been waiting...?}
\end{itemize}
\end{notebox}

\textbf{Scoring:} 1 point per correct question. Total: 8 points.

\newpage

% ============================================================
\section{Exercise 3: Affirmative, Negative, or Question? (10 points)}

\begin{answerbox}[Answers]
\begin{enumerate}[label=\arabic*.]
    \item (+) It \textbf{\textcolor{correct}{had been raining}}.
    \item (-) I \textbf{\textcolor{correct}{hadn't been working}} hard.
    \item (?) \textbf{\textcolor{correct}{Had you been waiting}} long when she finally called?
    \item (+) He \textbf{\textcolor{correct}{had been crying}}.
    \item (-) They \textbf{\textcolor{correct}{hadn't been paying}} attention.
    \item (?) How long \textbf{\textcolor{correct}{had she been learning}} Spanish before she moved to Madrid?
    \item (+) We \textbf{\textcolor{correct}{had been hiking}} all day.
    \item (-) The team \textbf{\textcolor{correct}{hadn't been practicing}} enough.
    \item (?) \textbf{\textcolor{correct}{Had he been drinking}} before the accident?
    \item (+) She \textbf{\textcolor{correct}{had been living}} there for years.
\end{enumerate}
\end{answerbox}

\textbf{Scoring:} 1 point per correct answer. Total: 10 points.

\vspace{0.5cm}

% ============================================================
\section{Exercise 4: Match the Cause and Effect (8 points)}

\begin{answerbox}[Answers]
\begin{center}
\begin{tabular}{|c|c|l|}
\hline
\textbf{Effect} & \textbf{Cause} & \textbf{Explanation} \\
\hline
1 & D & He was out of breath because he \textbf{had been running}. \\
\hline
2 & A & Her eyes were red because she \textbf{had been crying}. \\
\hline
3 & B & The streets were wet because it \textbf{had been raining}. \\
\hline
4 & C & I was very tired because I \textbf{had been working} all night. \\
\hline
5 & F & His hands were dirty because he \textbf{had been working} in the garden. \\
\hline
6 & G & She spoke perfect French because she \textbf{had been living} in Paris for 10 years. \\
\hline
7 & E & They were very hungry because they \textbf{had been traveling} all day without eating. \\
\hline
8 & H & The car broke down because it \textbf{had been making} strange noises for weeks. \\
\hline
\end{tabular}
\end{center}
\end{answerbox}

\begin{notebox}[Key Concept: Cause and Visible Result]
The Past Perfect Continuous is often used to explain WHY something was visible or true at a past moment:
\begin{itemize}
    \item \textbf{Visible result:} He was out of breath. (past state)
    \item \textbf{Cause/Activity:} He had been running. (continuous activity before)
\end{itemize}
\end{notebox}

\textbf{Scoring:} 1 point per correct match. Total: 8 points.

\newpage

% ============================================================
% MEDIUM SECTION ANSWERS
% ============================================================

\begin{center}
{\Large\textcolor{mediumyellow}{\textbf{— MEDIUM EXERCISES —}}}
\end{center}

\vspace{0.3cm}

% ============================================================
\section{Exercise 5: Past Perfect Continuous or Past Perfect Simple? (10 points)}

\begin{answerbox}[Answers]
\begin{enumerate}[label=\arabic*.]
    \item \textbf{\textcolor{correct}{had been waiting}} — Focus on duration (continuous)
    \item \textbf{\textcolor{correct}{had read}} — Focus on completed quantity (simple)
    \item \textbf{\textcolor{correct}{had been swimming}} — Focus on activity causing visible result (continuous)
    \item \textbf{\textcolor{correct}{had finished}} — Focus on completion (simple)
    \item \textbf{\textcolor{correct}{had been arguing}} — Focus on duration (continuous)
    \item \textbf{\textcolor{correct}{had never seen}} — Experience, not process (simple)
    \item \textbf{\textcolor{correct}{had been working}} — Continuous activity causing tiredness (continuous)
    \item \textbf{\textcolor{correct}{had already eaten}} — Completed action (simple)
    \item \textbf{\textcolor{correct}{had been playing}} — Activity causing visible result (continuous)
    \item \textbf{\textcolor{correct}{had been studying}} — Prolonged activity (continuous)
\end{enumerate}
\end{answerbox}

\begin{notebox}[When to Use Each Tense]
\textbf{Past Perfect Continuous:}
\begin{itemize}
    \item Emphasizes \textbf{duration/process}: ``for hours'', ``all day'', ``since morning''
    \item Explains \textbf{visible results}: tired, wet, dirty, out of breath
\end{itemize}

\textbf{Past Perfect Simple:}
\begin{itemize}
    \item Emphasizes \textbf{completion}: finished, done, completed
    \item Focuses on \textbf{quantity}: read 3 books, eaten 5 apples
    \item Used with \textbf{stative verbs}: know, believe, want, like (NOT continuous)
\end{itemize}
\end{notebox}

\textbf{Scoring:} 1 point per correct answer. Total: 10 points.

\newpage

% ============================================================
\section{Exercise 6: Complete the Story (16 points)}

\begin{answerbox}[Answers]
\begin{enumerate}[label=(\arabic*)]
    \item \textbf{\textcolor{correct}{woke up}} — Past Simple (main event)
    \item \textbf{\textcolor{correct}{realized}} — Past Simple (main event)
    \item \textbf{\textcolor{correct}{hadn't set}} — Past Perfect Simple (completed action before)
    \item \textbf{\textcolor{correct}{was}} — Past Simple (state)
    \item \textbf{\textcolor{correct}{had been trying}} — Past Perfect Continuous (repeated action over time)
    \item \textbf{\textcolor{correct}{answered}} — Past Simple (main event)
    \item \textbf{\textcolor{correct}{had been waiting}} — Past Perfect Continuous (duration)
    \item \textbf{\textcolor{correct}{had}} — Past Simple (scheduled event)
    \item \textbf{\textcolor{correct}{arrived}} — Past Simple (main event)
    \item \textbf{\textcolor{correct}{had already started}} — Past Perfect Simple (completed before)
    \item \textbf{\textcolor{correct}{was presenting}} — Past Continuous (action in progress)
    \item \textbf{\textcolor{correct}{had been preparing}} — Past Perfect Continuous (duration)
    \item \textbf{\textcolor{correct}{felt}} — Past Simple (emotion)
    \item \textbf{\textcolor{correct}{had been working}} — Past Perfect Continuous (duration)
    \item \textbf{\textcolor{correct}{got}} — Past Simple (main event)
    \item \textbf{\textcolor{correct}{went}} — Past Simple (main event)
\end{enumerate}
\end{answerbox}

\textbf{Scoring:} 1 point per correct answer. Total: 16 points.

\vspace{0.5cm}

% ============================================================
\section{Exercise 7: Multiple Choice (10 points)}

\begin{answerbox}[Answers]
\begin{enumerate}[label=\arabic*.]
    \item \textbf{\textcolor{correct}{A) had already escaped}} — Completed action (the thief was gone)
    \item \textbf{\textcolor{correct}{B) had been working}} — Continuous activity explaining exhaustion
    \item \textbf{\textcolor{correct}{A) had they been discussing}} — Duration question
    \item \textbf{\textcolor{correct}{B) had started}} — Completed event (NOT \textit{had been starting})
    \item \textbf{\textcolor{correct}{B) hadn't been sleeping}} — Continuous activity causing visible result
    \item \textbf{\textcolor{correct}{C) had been driving}} — Duration (``for three hours'')
    \item \textbf{\textcolor{correct}{B) had been raining}} — Continuous activity causing visible result
    \item \textbf{\textcolor{correct}{B) had been singing}} — Prolonged activity before another past event
    \item \textbf{\textcolor{correct}{A) had drunk}} — Focus on quantity (five cups)
    \item \textbf{\textcolor{correct}{B) had been thinking}} — Continuous process before that moment
\end{enumerate}
\end{answerbox}

\begin{notebox}[Common Trap]
\textbf{Some verbs are NOT normally used in continuous forms:}
\begin{itemize}
    \item \sout{had been starting} → \textbf{had started}
    \item \sout{had been knowing} → \textbf{had known}
    \item \sout{had been believing} → \textbf{had believed}
\end{itemize}
\end{notebox}

\textbf{Scoring:} 1 point per correct answer. Total: 10 points.

\newpage

% ============================================================
\section{Exercise 8: Time Expressions (8 points)}

\begin{answerbox}[Answers]
\begin{enumerate}[label=\arabic*.]
    \item I was exhausted because I had been studying \textbf{\textcolor{correct}{all night}}.
    \item When I finally saw her, she told me she had been waiting \textbf{\textcolor{correct}{for ages}}.
    \item The baby finally fell asleep. She had been crying \textbf{\textcolor{correct}{for two hours}}.
    \item He had been learning Japanese \textbf{\textcolor{correct}{since January}} before he moved to Tokyo.
    \item They were starving because they had been hiking \textbf{\textcolor{correct}{all day}} without eating.
    \item I couldn't sleep. I had been thinking about the problem \textbf{\textcolor{correct}{all night}}.
    \item She looked terrible. She had been feeling ill \textbf{\textcolor{correct}{for weeks}}.
    \item He got the job! He had been applying to companies \textbf{\textcolor{correct}{for ages / since morning}}.
\end{enumerate}
\end{answerbox}

\begin{notebox}[Time Expression Guide]
\begin{tabular}{|l|l|}
\hline
\textbf{FOR + duration} & for ages, for two hours, for weeks \\
\hline
\textbf{SINCE + starting point} & since morning, since January \\
\hline
\textbf{ALL + time period} & all day, all night \\
\hline
\end{tabular}
\end{notebox}

\textbf{Scoring:} 1 point per correct answer. Total: 8 points.

\textit{Note: Some answers may have acceptable alternatives (e.g., ``for ages'' or ``for weeks'' in \#8). Accept logical variations.}

\newpage

% ============================================================
% HARD SECTION ANSWERS
% ============================================================

\begin{center}
{\Large\textcolor{hardred}{\textbf{— HARD EXERCISES —}}}
\end{center}

\vspace{0.3cm}

% ============================================================
\section{Exercise 9: Error Correction (8 points)}

\begin{answerbox}[Answers]
\begin{enumerate}[label=\arabic*.]
    \item \textbf{Error:} ``had been work'' → missing -ing\\
    \textbf{\textcolor{correct}{Correction:}} She \textbf{had been working} all day, so she was very tired.
    
    \vspace{0.2cm}
    
    \item \textbf{Error:} ``have been waiting'' → wrong auxiliary (should be past)\\
    \textbf{\textcolor{correct}{Correction:}} They \textbf{had been waiting} for three hours when the train finally arrived.
    
    \vspace{0.2cm}
    
    \item \textbf{Error:} ``had been knowing'' → ``know'' is a stative verb\\
    \textbf{\textcolor{correct}{Correction:}} I \textbf{had known} her for years before we started dating.
    
    \vspace{0.2cm}
    
    \item \textbf{Error:} ``been wait'' → missing -ing\\
    \textbf{\textcolor{correct}{Correction:}} \textbf{Had you been waiting} long when I called?
    
    \vspace{0.2cm}
    
    \item \textbf{Error:} ``since five years'' → ``since'' needs a point in time, not duration\\
    \textbf{\textcolor{correct}{Correction:}} He had been living in London \textbf{for five years} before he moved.
    
    \vspace{0.2cm}
    
    \item \textbf{Error:} ``has been raining'' → wrong tense (should be past)\\
    \textbf{\textcolor{correct}{Correction:}} The ground was wet because it \textbf{had been raining}.
    
    \vspace{0.2cm}
    
    \item \textbf{Error:} ``hadn't been believed'' → ``believe'' is a stative verb\\
    \textbf{\textcolor{correct}{Correction:}} She \textbf{hadn't believed} in herself until she won the competition.
    
    \vspace{0.2cm}
    
    \item \textbf{Error:} ``had she been studied'' → mixing forms\\
    \textbf{\textcolor{correct}{Correction:}} How long \textbf{had she been studying} before she passed the exam?
\end{enumerate}
\end{answerbox}

\textbf{Scoring:} 1 point per correct correction. Total: 8 points.

\newpage

% ============================================================
\section{Exercise 10: Translation — Portuguese to English (8 points)}

\begin{answerbox}[Answers]
\begin{enumerate}[label=\arabic*.]
    \item Ele estava cansado porque tinha estado trabalhando a noite toda.\\
    \textbf{\textcolor{correct}{He was tired because he had been working all night.}}
    
    \vspace{0.3cm}
    
    \item Há quanto tempo você estava esperando quando ela chegou?\\
    \textbf{\textcolor{correct}{How long had you been waiting when she arrived?}}
    
    \vspace{0.3cm}
    
    \item Eles tinham estado morando em São Paulo por dez anos antes de se mudarem.\\
    \textbf{\textcolor{correct}{They had been living in São Paulo for ten years before they moved.}}
    
    \vspace{0.3cm}
    
    \item Estava chovendo o dia todo, então as ruas estavam molhadas.\\
    \textbf{\textcolor{correct}{It had been raining all day, so the streets were wet.}}
    
    \vspace{0.3cm}
    
    \item Ela não tinha estado se sentindo bem ultimamente.\\
    \textbf{\textcolor{correct}{She hadn't been feeling well lately.}}
    
    \vspace{0.3cm}
    
    \item Nós tínhamos estado estudando por horas quando decidimos fazer uma pausa.\\
    \textbf{\textcolor{correct}{We had been studying for hours when we decided to take a break.}}
    
    \vspace{0.3cm}
    
    \item Os olhos dele estavam vermelhos porque ele tinha estado chorando.\\
    \textbf{\textcolor{correct}{His eyes were red because he had been crying.}}
    
    \vspace{0.3cm}
    
    \item Você tinha estado praticando antes da competição?\\
    \textbf{\textcolor{correct}{Had you been practicing before the competition?}}
\end{enumerate}
\end{answerbox}

\textbf{Scoring:} 1 point per correct translation. Total: 8 points.

\textit{Note: Accept minor variations that are grammatically correct.}

\newpage

% ============================================================
\section{Exercise 11: Complete the Dialogue (13 points)}

\begin{answerbox}[Answers]
\begin{enumerate}[label=(\arabic*)]
    \item \textbf{\textcolor{correct}{applied}} — Past Simple (completed action)
    \item \textbf{\textcolor{correct}{had been working}} — Past Perfect Continuous (duration: ``for five years'')
    \item \textbf{\textcolor{correct}{did you decide}} — Past Simple (question about specific decision)
    \item \textbf{\textcolor{correct}{had been thinking}} — Past Perfect Continuous (process over time)
    \item \textbf{\textcolor{correct}{had been feeling}} — Past Perfect Continuous (ongoing state)
    \item \textbf{\textcolor{correct}{wasn't growing / hadn't been growing}} — Past Continuous or Past Perfect Continuous (both acceptable)
    \item \textbf{\textcolor{correct}{had you been working}} — Past Perfect Continuous (duration question)
    \item \textbf{\textcolor{correct}{had been managing / managed}} — Past Perfect Continuous or Past Simple (both acceptable)
    \item \textbf{\textcolor{correct}{left}} — Past Simple (completed action)
    \item \textbf{\textcolor{correct}{had been leading}} — Past Perfect Continuous (duration: ``for two years'')
    \item \textbf{\textcolor{correct}{Had you ever worked}} — Past Perfect Simple (experience question)
    \item \textbf{\textcolor{correct}{had collaborated}} — Past Perfect Simple (completed experience)
    \item \textbf{\textcolor{correct}{had been handling}} — Past Perfect Continuous (duration: ``for about three years'')
\end{enumerate}
\end{answerbox}

\textbf{Scoring:} 1 point per correct answer. Total: 13 points.

\vspace{0.5cm}

% ============================================================
\section{Exercise 12: Write Your Own Sentences (6 points)}

\begin{answerbox}[Sample Answers]
\begin{enumerate}[label=\arabic*.]
    \item \textbf{\textcolor{correct}{She was tired because she had been studying all night.}}\\
    (Or: ``He looked exhausted because he had been studying all night.'')
    
    \vspace{0.2cm}
    
    \item \textbf{\textcolor{correct}{How long had you been waiting when the bus arrived?}}\\
    (Or: ``How long had they been waiting before she called?'')
    
    \vspace{0.2cm}
    
    \item \textbf{\textcolor{correct}{I had been looking for work for months before I got this job.}}\\
    (Or: ``She had been looking for work for months before she found a position.'')
    
    \vspace{0.2cm}
    
    \item \textbf{\textcolor{correct}{He hadn't been listening, so he didn't understand the instructions.}}\\
    (Or: ``They hadn't been listening during the meeting.'')
    
    \vspace{0.2cm}
    
    \item \textbf{\textcolor{correct}{The streets were wet because it had been raining all morning.}}
    
    \vspace{0.2cm}
    
    \item \textbf{\textcolor{correct}{She had been practicing the piano for years before she won the competition.}}\\
    (Or: ``He had been practicing for years before he became a professional.'')
\end{enumerate}
\end{answerbox}

\textbf{Scoring:} 1 point per grammatically correct sentence using Past Perfect Continuous appropriately. Total: 6 points.

\textit{Accept any grammatically correct sentence that properly uses the Past Perfect Continuous with the given prompts.}

\newpage

% ============================================================
% REAL-LIFE & EXAM SECTION ANSWERS
% ============================================================

\begin{center}
{\Large\textcolor{blue!70!black}{\textbf{--- REAL-LIFE \& EXAM PRACTICE ---}}}
\end{center}

\vspace{0.3cm}

% ============================================================
\section{Exercise 13: Micro-Dialogues (8 points)}

\begin{answerbox}[Answers]
\begin{enumerate}[label=\arabic*.]
    \item B: ``About two hours. I \textbf{\textcolor{correct}{had been lifting}} weights before you arrived, and then I did cardio.''
    
    \item B: ``Yes, I \textbf{\textcolor{correct}{had been waiting}} for 30 minutes! I almost left.''
    
    \item B: ``She \textbf{\textcolor{correct}{had been working}} on that report all week. The deadline is tomorrow.''
    
    \item B: ``I \textbf{\textcolor{correct}{had been cooking}} all afternoon for the dinner party tonight.''
    
    \item B: ``I \textbf{\textcolor{correct}{hadn't been sleeping}} well lately. Too much stress at work.''
    
    \item B: ``I \textbf{\textcolor{correct}{had been studying}} for months before I took it. Hard work pays off!''
    
    \item B: ``Well, I \textbf{\textcolor{correct}{had been exercising}} regularly and eating healthier.''\\
    \textit{(Note: ``have been exercising'' is also acceptable here, depending on whether the action continues)}
    
    \item B: ``It started yesterday, but I \textbf{\textcolor{correct}{had been feeling}} tired for weeks before that.''
\end{enumerate}
\end{answerbox}

\begin{notebox}[Why Past Perfect Continuous Here?]
These dialogues describe activities that were \textbf{ongoing before another past event}:
\begin{itemize}
    \item ``before you arrived'' — past reference point
    \item ``for 30 minutes'' — duration emphasis
    \item ``all week / all afternoon'' — continuous activity
\end{itemize}
\end{notebox}

\textbf{Scoring:} 1 point per correct answer. Total: 8 points.

\newpage

% ============================================================
\section{Exercise 14: Real-Life Context (15 points)}

\begin{answerbox}[Scenario A: Job Interview Email]
\begin{enumerate}[label=(\arabic*)]
    \item before I \textbf{\textcolor{correct}{joined}} my current company — \textit{(Past Simple: completed action)}
    \item I \textbf{\textcolor{correct}{had been working}} as a freelance consultant — \textit{(PPC: ongoing before)}
    \item I \textbf{\textcolor{correct}{had been developing}} the skills — \textit{(PPC: process over time)}
    \item I \textbf{\textcolor{correct}{had been thinking}} about transitioning — \textit{(PPC: mental process)}
\end{enumerate}
\end{answerbox}

\begin{answerbox}[Scenario B: Academic Report]
\begin{enumerate}[resume, label=(\arabic*)]
    \item Before this study \textbf{\textcolor{correct}{began}} — \textit{(Past Simple: point in time)}
    \item researchers \textbf{\textcolor{correct}{had been investigating}} — \textit{(PPC: ongoing research)}
    \item Previous studies \textbf{\textcolor{correct}{had shown}} — \textit{(PP Simple: completed findings)}
    \item no conclusive data \textbf{\textcolor{correct}{had existed}} — \textit{(PP Simple: state, not process)}
    \item Our team \textbf{\textcolor{correct}{had been collecting}} data — \textit{(PPC: ongoing process)}
    \item before we \textbf{\textcolor{correct}{published}} these findings — \textit{(Past Simple: completed)}
\end{enumerate}
\end{answerbox}

\begin{answerbox}[Scenario C: News Report]
\begin{enumerate}[resume, label=(\arabic*)]
    \item The company \textbf{\textcolor{correct}{had been planning}} this expansion — \textit{(PPC: ongoing)}
    \item ``We \textbf{\textcolor{correct}{had been looking}} at several locations'' — \textit{(PPC: process)}
    \item before we \textbf{\textcolor{correct}{chose}} São Paulo — \textit{(Past Simple: decision point)}
    \item The company \textbf{\textcolor{correct}{has been growing}} — \textit{(Present Perf. Cont.: up to now)}
    \item and \textbf{\textcolor{correct}{has hired}} over 200 employees — \textit{(Present Perf.: result)}
\end{enumerate}
\end{answerbox}

\begin{notebox}[Professional Writing Tips]
\textbf{In formal writing:}
\begin{itemize}
    \item Use PPC to describe \textbf{background context}
    \item Combine with Past Simple for \textbf{specific events}
    \item In news: switch to Present Perfect for \textbf{recent/ongoing news}
\end{itemize}
\end{notebox}

\textbf{Scoring:} 1 point per correct answer. Total: 15 points.

\newpage

% ============================================================
\section{Exercise 15: Common Pitfalls for Brazilians (8 points)}

\begin{answerbox}[Corrections]
\begin{enumerate}[label=\arabic*.]
    \item \textbf{\textcolor{correct}{had been working}} (not ``have been working'')\\
    \textit{Why? The reference point (``when you called'') is in the past.}
    
    \item \textbf{\textcolor{correct}{for}} three hours (not ``since'')\\
    \textit{Why? ``Three hours'' is a duration, not a starting point.}
    
    \item \textbf{\textcolor{correct}{had known}} him (not ``had been knowing'')\\
    \textit{Why? ``Know'' is a stative verb — no continuous form.}
    
    \item \textbf{\textcolor{correct}{How long had you been waiting?}}\\
    \textit{Why? ``Had'' must come before the subject in questions.}
    
    \item \textbf{\textcolor{correct}{He had been working}} all day.\\
    \textit{Why? Don't add ``was being'' — it's redundant.}
    
    \item \textbf{\textcolor{correct}{They had been waiting}} for hours.\\
    \textit{Why? ``Been'' is essential to the structure.}
    
    \item \textbf{\textcolor{correct}{By 6 PM, I had finished}} my homework.\\
    \textit{Why? Completed actions use Past Perfect Simple.}
    
    \item \textbf{\textcolor{correct}{She had been working}} hard all week.\\
    \textit{Why? Always use verb-ing after ``had been.''}
\end{enumerate}
\end{answerbox}

\begin{notebox}[Common Portuguese Interference]
\textbf{Watch out for these patterns:}
\begin{itemize}
    \item PT: ``Eu \textit{estava trabalhando}...'' $\neq$ EN: ``I \textit{was being working}''
    \item PT: ``Desde 3 horas'' $\rightarrow$ EN: ``\textbf{For} 3 hours'' (duration)
    \item PT: ``Como você tem estado?'' $\rightarrow$ Context matters for tense choice!
\end{itemize}
\end{notebox}

\textbf{Scoring:} 1 point per correct correction. Total: 8 points.

\newpage

% ============================================================
\section{Exercise 16: IELTS Practice (Sample Answers)}

\begin{notebox}[Assessment Criteria]
\textbf{IELTS evaluates:} Grammatical Range \& Accuracy, Lexical Resource, Coherence \& Cohesion.\\
Using Past Perfect Continuous correctly demonstrates \textbf{grammatical range}.
\end{notebox}

\begin{answerbox}[Part A: Speaking Part 2 — Sample Response]
``I'd like to talk about passing my driving test, which I finally achieved last year.

I \textbf{had been working} towards this goal for almost two years. Before I passed, I \textbf{had been taking} lessons every week and practicing with my father on weekends.

The main difficulty was that I \textbf{had been struggling} with parallel parking for months. I \textbf{had been failing} the practical test twice before, which was very frustrating.

When I finally passed, I felt incredibly relieved. I \textbf{had been dreaming} about this moment for so long that I almost cried with happiness.''
\end{answerbox}

\vspace{0.3cm}

\begin{answerbox}[Part B: Speaking Part 3 — Sample Answers]
\begin{enumerate}[label=\arabic*.]
    \item ``In Brazil, environmental activists and scientists \textbf{had been discussing} issues like deforestation and pollution for decades before the government introduced significant policies. NGOs \textbf{had been campaigning} for years before real changes happened.''
    
    \item ``Before online learning became mainstream, educational institutions \textbf{had been experimenting} with digital tools. Some universities \textbf{had been offering} distance learning programs, but they weren't widely accepted.''
    
    \item ``I think some people \textbf{had been expecting} technological changes, especially those who work in tech. However, most people \textbf{hadn't been anticipating} how quickly everything would transform.''
\end{enumerate}
\end{answerbox}

\vspace{0.3cm}

\begin{answerbox}[Part C: Writing Task 2 — Sample Introduction]
``Before the digital revolution transformed our daily lives, humans \textbf{had been living} relatively simply for generations. People \textbf{had been communicating} through letters and landline phones, and they \textbf{had been managing} their work without constant connectivity. This essay will argue that while technology has added certain complexities, it has ultimately improved our quality of life.''
\end{answerbox}

\textbf{Scoring:} Self-assessment based on correct PPC usage. Award up to 8 points.

\newpage

% ============================================================
\section{Exercise 17: Cambridge Exam Practice (12 points)}

\begin{answerbox}[Part A: Key Word Transformations]
\begin{enumerate}[label=\arabic*.]
    \item At 10 AM, she \textbf{\textcolor{correct}{had been studying for}} four hours.
    
    \item It \textbf{\textcolor{correct}{had been raining}} all morning, so the streets were wet.
    
    \item They \textbf{\textcolor{correct}{had been negotiating}} for three months before they agreed.
    
    \item He \textbf{\textcolor{correct}{had been practicing}} for years before he won at 18.
    
    \item I \textbf{\textcolor{correct}{had been waiting}} for two hours when she arrived.
\end{enumerate}
\end{answerbox}

\begin{notebox}[Cambridge Key Word Transformations]
\textbf{Tips for success:}
\begin{itemize}
    \item Count words carefully (contractions = 2 words)
    \item Don't change the key word
    \item Keep the \textbf{exact meaning} — don't add or lose information
\end{itemize}
\end{notebox}

\vspace{0.3cm}

\begin{answerbox}[Part B: Open Cloze]
\begin{enumerate}[label=(\arabic*)]
    \item \textbf{\textcolor{correct}{had}}
    \item \textbf{\textcolor{correct}{had}}
    \item \textbf{\textcolor{correct}{had}}
    \item \textbf{\textcolor{correct}{had}}
    \item \textbf{\textcolor{correct}{had}}
    \item \textbf{\textcolor{correct}{have}} (or ``had'')
    \item \textbf{\textcolor{correct}{had}}
\end{enumerate}
\end{answerbox}

\textbf{Scoring:} Key Word Transformations: 1 point each (5 points). Open Cloze: 1 point each (7 points). Total: 12 points.

\newpage

% ============================================================
\section{Exercise 18: TOEFL Practice (Sample Answers)}

\begin{answerbox}[Part A: Speaking Task — Sample Response]
``The student supports the library's decision to extend its hours for two main reasons.

First, she explains that she \textbf{had been struggling} to find quiet study spaces before this change. The library closes early, so she \textbf{had been having} difficulty finding a peaceful place to study at night.

Second, she mentions that she \textbf{had been going} to coffee shops to study late, which was both expensive and distracting. With longer library hours, she won't need to spend money or deal with noise anymore.''
\end{answerbox}

\begin{notebox}[TOEFL Speaking Tips]
\textbf{Integrated Tasks:}
\begin{itemize}
    \item Clearly state the student's opinion/preference
    \item Use PPC to describe \textbf{previous situations}
    \item Connect reasons logically with transition words
\end{itemize}
\end{notebox}

\vspace{0.3cm}

\begin{answerbox}[Part B: Integrated Writing — Sample Response]
``The reading and lecture present contrasting perspectives on remote work. While the reading states that remote work has become common with stable productivity, the lecture provides historical context.

According to the professor, before the pandemic only 5\% of workers \textbf{had been working} remotely. Furthermore, many companies \textbf{had been resisting} flexible arrangements for years. This context suggests that the recent shift to remote work represents a dramatic departure from previous workplace norms, rather than a gradual evolution.''
\end{answerbox}

\textbf{Scoring:} Self-assessment based on correct PPC usage and coherent argument. Award up to 4 points.

\newpage

% ============================================================
% FINAL SCORING
% ============================================================

\begin{center}
\rule{0.8\textwidth}{0.4pt}\\[0.5cm]
{\Large\bfseries Total Score Summary}\\[0.5cm]

\begin{tabular}{|l|c|c|}
\hline
\textbf{Exercise} & \textbf{Points} & \textbf{Your Score} \\
\hline
\multicolumn{3}{|c|}{\textbf{\textcolor{easygreen}{EASY SECTION}}} \\
\hline
1. Form the Past Perfect Continuous & 10 & \rule{1.5cm}{0.4pt} \\
2. Make Questions & 8 & \rule{1.5cm}{0.4pt} \\
3. Affirmative, Negative, or Question & 10 & \rule{1.5cm}{0.4pt} \\
4. Match Cause and Effect & 8 & \rule{1.5cm}{0.4pt} \\
\hline
\textit{Easy Subtotal} & \textit{36} & \rule{1.5cm}{0.4pt} \\
\hline
\multicolumn{3}{|c|}{\textbf{\textcolor{mediumyellow}{MEDIUM SECTION}}} \\
\hline
5. Past Perfect Continuous or Simple? & 10 & \rule{1.5cm}{0.4pt} \\
6. Complete the Story & 16 & \rule{1.5cm}{0.4pt} \\
7. Multiple Choice & 10 & \rule{1.5cm}{0.4pt} \\
8. Time Expressions & 8 & \rule{1.5cm}{0.4pt} \\
\hline
\textit{Medium Subtotal} & \textit{44} & \rule{1.5cm}{0.4pt} \\
\hline
\multicolumn{3}{|c|}{\textbf{\textcolor{hardred}{HARD SECTION}}} \\
\hline
9. Error Correction & 8 & \rule{1.5cm}{0.4pt} \\
10. Translation (PT → EN) & 8 & \rule{1.5cm}{0.4pt} \\
11. Complete the Dialogue & 13 & \rule{1.5cm}{0.4pt} \\
12. Write Your Own Sentences & 6 & \rule{1.5cm}{0.4pt} \\
\hline
\textit{Hard Subtotal} & \textit{35} & \rule{1.5cm}{0.4pt} \\
\hline
\multicolumn{3}{|c|}{\textbf{\textcolor{blue!70!black}{REAL-LIFE \& EXAM}}} \\
\hline
13. Micro-Dialogues & 8 & \rule{1.5cm}{0.4pt} \\
14. Real-Life Context & 15 & \rule{1.5cm}{0.4pt} \\
15. Common Pitfalls & 8 & \rule{1.5cm}{0.4pt} \\
16. IELTS Practice & 8 & \rule{1.5cm}{0.4pt} \\
17. Cambridge Practice & 12 & \rule{1.5cm}{0.4pt} \\
18. TOEFL Practice & 4 & \rule{1.5cm}{0.4pt} \\
\hline
\textit{Exam Subtotal} & \textit{55} & \rule{1.5cm}{0.4pt} \\
\hline
\textbf{GRAND TOTAL} & \textbf{170} & \rule{1.5cm}{0.4pt} \\
\hline
\end{tabular}

\vspace{1cm}

\begin{tabular}{|l|l|}
\hline
\textbf{Score Range} & \textbf{Level} \\
\hline
153-170 (90\%+) & Excellent! Ready for advanced grammar \\
\hline
136-152 (80-89\%) & Very Good! Minor review needed \\
\hline
102-135 (60-79\%) & Good, but review weak areas \\
\hline
Below 102 (<60\%) & More practice recommended \\
\hline
\end{tabular}

\vspace{0.5cm}

\rule{0.8\textwidth}{0.4pt}\\[0.3cm]
{\large\textit{Great work completing these exercises!}}\\
{\textit{Review any mistakes and practice the Past Perfect Continuous in your own writing.}}
\end{center}

\end{document}
